\textbf{Microbiota predict \emph{Clostridium difficile} severity in germ-free mice colonized with human feces}

### 2021-11-30 submission/manuscript.Rmd
revisions based on ST and SL feedback

# The gut microbiota potentiates *Clostridiodes difficile* infection severity.

@@

Nicholas A. Lesniak$^1$, Alyxandria M. Schubert$^1$, Kaitlyn Flynn$^1$, Hamide Sinani$^1$, Kathyrn A. Eaton$^1$, Ingrid L. Bergin$^2$, Patrick D. Schloss$^{1,\dagger}$

@@

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu  
1\. Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI  
2\. Unit for Laboratory Animal Medicine, University of Michigan, Ann Arbor, MI

@@

*Clostridioides difficile* infection has become more common and severe in the last few decades. Patient age, white blood cell count, creatinine levels as well as *C. difficile* ribotype and the presence of toxin genes have been associated with disease severity. However, models based on those data to identify patients with CDI that will develop severe disease have not been robust enough to be broadly applied. These models are input from our understanding of *C. difficile* interactions with the gut environment. The gut bacteria are a key determinant in *C. difficile* infection. CDI is dependent on perturbations in the bacterial community in order to colonize the gut. The gut microbiota also impairs *C. difficile* colonization through bile acid metabolism, nutrient consumption and bacteriocin production. However, it is unclear if the gut bacteria affect the disease severity resulting from that colonization. Here we have shown gut bacterial communities contribute to disease severity variation. We created diverse gut communities by colonizing germ-free mice with human fecal communities. They were then infected with *C. difficile* which resulted in differed disease severities. The severity grouped by the human fecal community they received. Generally, facultative anaerobes with pathogenic potential, such as Escherichia, Helicobacter  and Klebsiella, were associated with more severe disease. Bacterial groups associated with dietary fiber degradation, such as Coprobacillus, were associated with reduced disease severity. Lastly, we showed communities that resulted in either low or high histopathologic scores or severe disease had the same outcome when infected with a different *C. difficile* isolate.

@@

Clostridioides difficile infection (CDI) causes a range of disease from asymptomatic to mild diarrhea to severe outcomes of recurrent infections and even death. Models that predict severity and treatment decisions are based on clinical factors and *C. difficile* capabilities. But currently the gut microbiome, the primary protector against CDI, has not been accounted for its effect on CDI disease severity. We demonstrated variation in the microbiome of mice colonized with human feces was sufficient to result in a range of disease severities. Our results show groups of bacteria contribute to the development of disease in *C. difficile* infections. Gut bacterial community data from patients with CDI could improve our ability to identify patients more at risk for severe disease and improve interventions which treat the gut bacteria to reduce host damage.

@@

*Clostridioides difficile* infections (CDI) have been increasing in incidence and severity since *C. difficile* was first identified as a pathogen. It was first identified in 1931 in a healthy fecal community [CITATION]. But then in the late 1970s,  following the introduction of broad-spectrum antibiotics, *C. difficile* was identified as the cause of antibiotic-associated psuedomembranous colitis [CITATION]. Beginning in the early 2000s, CDI became more problematic with increasing incidences and severity [CITATION]. Disease severities can range from asymptomatic to mild diarrhea to toxic megacolon and death. This severity has been associated with characteristics of *C. difficile* ribotypes and host factors. Over time more pathogenic ribotypes have emerged, such as ribotype 027 in the early 2000s. The presence of the genes for Toxin A/B or fluoroquinolon resistence has been associated with more severe disease [CITATION]. Binary toxin, while alone insufficient to cause damage on it own, when in the presence of Toxin A/B has been associated with increased virulence [CITATION]. Mutations in tcd C, the negative regulator of of toxin transcription, also increased the virulence of *C. difficile* [CITATION]. In addition to the pathogenic potential of the infecting strain, the response by the host also affects the disease severity. Delayed or reduced production of antibodies IgG and IgA can result in more severe disease [CITATION]. Increased neutrophil inflitration can lead to worse outcomes [CITATION]. In addition to the immune response, increased density of the toxin receptor on the host cells can increase the severity of the infection [CITATION]. While there is a thorough understanding of many of the mechanisms driving in CDI incidence and disease severity, there still is not effective tools to reduce the risk of severe disease.

Models have been developed to identify patients at risk for severe CDI but have not been robust for broad application [@Dieterle2020, @Butt2013, @Perry2021, @vanBeurden2017]. These models utilize patient related data, such as patient age, white blood cell count, serum albumin levels, and creatinine levels, to predict disease severity. In their initial publication, these models perform well enough for publication. However when Perry et al applied many of the most current models to a multi-center external validation, their performance suffered and often had more false positives than true positives [@Perry2021]. Thus, there seems to be a missing factor contributing to disease severity.

Missing from these models are the gut bacteria which also interact with *C. difficile* during the course of the infection. The indigenous bacteria of a healthy intestinal community provide a protective barrier preventing *C. difficile* from infecting the gut. Only after this community is disrupted from a range of pertubations, such as many classes of antibiotics, medications, or even dietary changes, can *C. difficile* infect the intestine. Once established, the gut bacteria can either promote or inhibit *C. difficile* through producing molecules or modifying the environment [Abbas2020]. Bile acids metabolized by the gut bacteria can inhibit *C. difficile* growth and affect toxin production [@Sorg2008, @Thanissery2017]. Bacteria in the gut also can compete more directly with *C. difficile* through antibiotic production or nutrient consumption [@Aguirre2021, @Kang2019]. To eliminate CDI, the primary treatment is a course of antibiotics. Patients who become reinfected and recalitrant to antibiotics, a fecal microbiota transplantion often is sufficient to restore the gut community to eliminate *C. difficile*. While there has been many studies to investigate the interaction between the gut bacteria and *C. difficile* during the infection, it has not been demostrated the bacterial community can modulate the disease severity of CDI. Recently, Tomkovich et al showed the variation in the bacterial communtiies between mice from different colonies and vendors was sufficient to cause differences in the temporal dynamics of the *C. difficile* infection [@Tomkovich2020]. Since the gut bacteria interact with *C. difficile* throughout the infection and community difference resulted in different colonization dynamics, we hypothesized that the gut bacteria contribute to the variation in CDI disease severity. Here, we colonized germ-free C57BL/6 mice with human fecal communities to test if the variablity in the gut bacteria can produce and explain variation in the disease severity caused by a single *C. difficile* isolate.

@@

**Germ-free mice colonized with human fecal commmunities have diverse and unique community structures.** Based on our previous observation mice with variation in their microbiota showed differences in colonization of *C. difficile* [@Tomkovich2020], we sought to explore the effect communties with greater variation had on *C. difficile* challenge. To produce communities with greater variation than mouse colonies with their indigenous communties, we used human fecal communities to establish a community in the intestine of germ-free mice. We inoculated a cage of germ-free C57BL/6 mice (2-4 mice per cage) with homogenized feces from one of 15 different human donors via oral gavage. The fecal community was allowed to colonize and stabilize for two weeks post inoculation [CITATION?]. We then assessed the gut bacterial communities by sequencing the V4 region of the 16S rRNA gene extracted from fecal pellets (Figure 1A). The communities established in the mice group most closely to other mice recieving the same human fecal donor community than to either their respective donor or any mice who received stool from a different donor (Figure 1B). Most of the communities were primairily composed of populations of *Clostridia*, *Bacteroidia*, *Erysipelotrichia*, *Bacilli*, and *Gammaproteobacteria*. However, each group of mice harbored unique combination of class abunadances in their gut bacterial communities.
*Any alpha? compare to murine variation? Number of unique otus?*

***C. difficile* is able to colonize mice without perturbation.** Our goal was to test the effect of the variation in the microbiota on *C. difficile* infection. A typical mouse model of CDI requires pretreatment with antibiotics such as clindamycin to become susceptible to *C. difficile* colonization [CITATION]. However, we wanted to avoid modifying the communties with an antibiotic to maintain their community structures and differences. Since these communties came from human donors which may have become susceptible through their own exposures, we decided to test if *C. difficile* was able to colonize these communities without any perturbation. After two weeks post incoluation with human fecal communities, the mice were then challenged with 10^**4?** *C. difficile* ribotype 027 isolate 431 spores (**Add schematic?**) [@Carlson2013]. The mice were followed for 10 days post challenge and their stool was collected and plated for *C. difficile* CFU to detect colonization level. We hypothesized that *C. difficile* would most likely be able to colonize the mice with communities that came from donors which had a history of antibiotic use, one of the primary risk factors for developing CDI (TABLE, CITATION). Suprisingly, communities from all donors were colonized (Figure 2). The only two mice that were able to resist *C. difficile* colonization both recieved their community from the same donor. Therefore, the transplanted human fecal communties were susceptible to *C. difficile* colonization without the need for any perturbation.

**Infection seveity varies by initial community structure.** After challenging the mice with *C. difficile*, we looked at the effect of the community variation on the outcome from the *C. difficile* infection. We followed the mice for signs of disease as well as *C. difficile* colonization and toxin production for 10 days post challenge. Some mice were not initally colonized on the day after *C. difficile* challenge but became colonized with a few days post challenge (**SUPP FIG?**). A subset of mice, all that recieved their comunity from one of 6 donors, suffered from severe disease, which became highly colonized one day post challenge and moribund within 3 days post challenge. The remaining mice, except for one cage, maintained *C. difficile* colonization through the end of the experiment (Figure 2). After the  mice were humanely euthanised at the endpoint, day 10 post challenge or earlier if they became moribund, their cecal tissue was collected and scored for histological signs of disease and fecal samples were assayed for toxin (Figure 3). Overall, there was greater toxin activity detected in the stool of the moribund mice (*P* = 0.003). However, when looking at each group of mice, we saw there was a range in toxin activity for both the mice with non-severe disease and severe disease (Figure 3A). Some donor groups that maintained colonization without severe disease had similarlly high activity levels as was detected in the mice with severe disease. Additionally, some mice with severe disease did not have toxin activity detected in their stool. These oberserved variations in toxin activity and relation to disease severity have been previously characterized [@Akerlund2006, @Vitucci2020]. Next we examined the cecal tissue for histopathological damage. For mice with severe disease, we observed high levels of epithelial damage, tissue edema and inflammation (Figure SXXXX), similar to previously reported histopathologic findings for *C. difficile* ribotype 027 [CITATION, @Cowardin2016]. As observed with toxin activity, the mice with severe disease had higher histopathologic scores than the mice with non-severe disease (*P* = 3.0e-9). However, unlike the toxin activity, all moribund mice had consistently high histopathologic scores. The mice which maintained persistent population size, we saw a range in tissue damage from no signs of disease to nearly the same level as the mice with severe disease, which grouped by community donor. Together, the toxin activity, histopathologic score, and moribundity showed variation across the donors but were largely consistent within each group of mice recieving the same human donor fecal community.

**Populations of the microbial communities explain variation in CDI disease severity.** Since the disease charateristics grouped by human donor fecal community, we next investigated the community composition for its ability to explain the variation in disease severity. First, we tested for associations between taxonomic groups and the outcomes of the *C. difficile* infection. We used the linear discriminant analysis (LDA) effect size (LEfSe) method to identify differences in the genera of the initial bacterial communities between the severity level and toxin production. We split the mice into groups by severity level based on their if they became moribund (severe), or if their clinical score was above 5 (moderate) or below 5 (mild). This analysis revealed 20 genera that were significantly different by the disease severity (Figure 4A). The presence of populations of *Turicibacter*, *Streptococcus*, *Staphylococcus*, *Pseudomonas*, *Phocaeicola*, *Parabacteroides*, *Bacteroides*, and *Escherichia/Shigella* were detected at higher levels in the mice that developed severe disease. Populations of *Anaerotignum*, *Coprobacillus*, *Enterocloster*, and *Murimonas* were increased in the mice that would experience mild disease. There were many genera that did not have a correlation of relative abundance and disease severity, such as the populations of *Monoglobus* and *Hungatella*, which were decreased or increased, respectively, in only the group of mice that would develop moderate disease. To understand these cases better we used LEfSe to identify the genera differentiating the communities that we detected toxin in from those that we did not (Figure 4B). There were many genera the were associated with toxin production as we would expect from their association with severe disease, such as populations of *Escherichia/Shigella* and *Bacteroides*. Likewise, there were genera such as *Anaerotignum*, *Enterocloster*, and *Murimonas* that were higher in communities that had mild disease and no toxin was produced. However, communities without a linear trend between relative abundance and disease severity can be better understood by their association with toxin production. Populations of *Hungatella* were increased in the group with moderate disease but not with the production of toxin. Lastly, we looked for associations between the histopathologic score and the populations of genera at that same time (Figure 4C). The populatons of *Bacteroides* at the end matched the trend observed in the initial community, increased population correlated with increased histopathologic score. Populations of *Klebsiella* and *Prevotellaceae* positively correlated with clinic score and were increased in the group that produced toxin. These tests have identified which individual populations of bacteria that associated with each disease severity.

We next wanted to understand how the populations in the context of each other lead to the different disease severities. We trained random forest models with the intital bacterial community relative abundance data at each taxonomic rank to predict toxin production, severe disease or final histopathologic score. Overall for predicting toxin production, microbial populations aggregated by their phylum level classificiation performed as well as models using lower taxonomic ranks (AUROC = 0.83, Figure S2). In the initial community populations of *Verrucomicrobia*, *Campilobacterota* and *Proteobacteria* contributed the most to the correct prediction of toxin production (Figure 5A). *C. difficile* was more likely to produce toxin when the initial community had smaller populations of *Verrucomicrobia* and *Campilobacterota* and had larger popluations of *Proteobacteria*. Next, we assessed the ability of the community to predict severe disease. Class rank classification was sufficient to predict if the mouse would succumb to the infection before the end of the experiment (AUROC = 0.91, Figure S2). Groups of bacteria belonging to *Bacilli*, *Firmicutes*, and *Erysipelotrichia* contributed the most to the correct prediction. Larger populations of *Bacilli* and *Firmicutes* and reduced populations of *Erysipelotrichia* were more likely to result in moribundity (Figure 5B). Aside from *Erysipelotrichia*, the only other population decreased in moribund mice was *Clostridia*, all other features had increased populations in the moribund mice. Lastly, the initial community was able to predict if the mice would have a histopathologic score above or below the median score (median histopathologic score = 5) with the genera relative abundances (AUROC = 0.99, Figure S2). There were no genera with a significantly greater effect on the model performance than any others, indicating the model is reliant on many genera for the correct prediction. The model used some of the genera identified in the LEfSe analysis, such as populations of *Coprobacillus*, *Anaerostipes*, and *Hungatella*. Communities with greater populations of *Hungatella*, *Eggerthella*, *Bifidobacterium*, *Duncaniella* and *Neisseria* were more likely to have greater histopathologic scores.

**Disease severity consistent by donor community across strains/isolates.** We used a single *C. difficile* isolate, ribotype 027 isolate 431 to characterize the range of disease severity and the features of the bacterial community associated with severity [@Carlson2013]. Since we had used the same mice and *C. difficile* isolate in our first set of experiemnts, we next wanted to test if the effect of the community would apply to other *C. difficile* isolates. We selected three communities based on the result from our analysis thus far to select one community we would expect to result in a low clinical score (< 5), a high clinical score (> 5), and one with a high clinical score which becomes moribund. Using genus level data to select the communities, we selected communities based on the relative abundance patterns of *Sellimonas* and *Anaerotignum* (higher abundance in mice with a low clinical score), *Lachnospiracea* (lower abundance in moribund mice), *Hungatella* and *Eisenbergiella* (higher abundance in moderate mice and lower in moribund mice), Clostridium sensu stricto (higher in moribund mice by both LEfSe analysis and RF model) and *Enterocloster* (lower in moribund mice and a negative linear trend with disease severity by LEfSe analysis) (Figure S3). With this set of human fecal communities selected, we inoculated mice via oral gavage of the human fecal slurry and allowed two weeks for the a community to establish. We then challenged the mice with a different *C. difficile* [@Carlson2013]. The disease severity for this set of experiments closely matched that in our first set of experiments with *C. difficile* isolate 431 (Figure 6).

@@

Challenging mice colonized with different human fecal communities with a single *C. difficile* isolate allowed us to identify the effect of microbiome variation on *C. difficile* infection disease severity. Our LEfSe analyses and random forest models revealed the relationships between the bacterial community members and the degree of disease severity. We used the relative abundance patterns of community members with strong associations with disease severity to select communities to test if the observed disease severity is conserved when other *C. difficile* isolates are used. The disease outcome from the *C. difficile* chalenge with the new isolate matched the expected disease severities. Overall, these results have demonstrated that the bacterial community is capable of modulating the disease severity of *C. difficile* infection.

CDI severity can range from no symptoms, mild symptoms such as diarrhea, or severe symptoms such as toxic megacolon or death. Physicians use prediction tools to identify patients most at risk of developing a severe infection using white blood cell counts, serum albumin level, or serum creatine level [@Lungulescu2011, @McDonald2018, @Zar2007]. Those levels are being driven by the activities in the intestine [@diMasi2018]. Research into the drivers of this variation have revealed factors that make *C. difficile* more virulent. Strains are categorized for their virulence by the presence and production toxin A and toxin B and the prevelance in outbreaks, such as ribotype 027 and 078 [@AbernathyClose2020, @Theriot2011, @Goorhuis2008, @OConnor2009, @Chen2008]. However, other studies have shown that virulence is not necessarily linked with toxin production **@Roa2015** or the strain **@Walk2012**. Furthermore, there is variation in the genome, growth rate, sporulation, germination, and toxin production in different isolates of a strain [@Carlson2013, @He2010]. This variation may help explain why severe CDI prediction tools often miss identifying many patients with CDI that will develop severe disease [@Dieterle2020, @Butt2013, @Perry2021, @vanBeurden2017]. Therefore, it is necessary to gain a full understanding of all factors contributing to disease variation to improve our ability to predict severity.

>>>>

# The gut bacterial community potentiates *Clostridiodes difficile* infection severity.

@@

Nicholas A. Lesniak$^1$, Alyxandria M. Schubert$^1$, Kaitlyn Flynn$^1$, Jhansi L Leslie$^1$, Hamide Sinani$^1$, Ingrid L. Bergin$^3$, Vincent B Young$^2$, Patrick D. Schloss$^{1,\dagger}$

@@

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu  
1\. Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI  
2\. Division of Infectious Diseases, Department of Internal Medicine, University of Michigan Medical School, Ann Arbor, MI
3\. Unit for Laboratory Animal Medicine, University of Michigan, Ann Arbor, MI

@@

*Clostridioides difficile* infection (CDI) has become more common and severe in the last few decades. Patient age, white blood cell count, creatinine levels as well as *C. difficile* ribotype and the presence of toxin genes have been associated with disease severity. However, current models of CDI disease severity that incorporate these factors are not robust enough to be broadly applied. These models are missing input from the gut environment. CDI is dependent on perturbations of the gut bacterial community in order to colonize the gut. The gut microbiota also impairs *C. difficile* colonization through bile acid metabolism, nutrient consumption and bacteriocin production. However, it is unclear if the gut bacteria affect CDI disease severity. Here we demonstrate that gut bacterial communities contribute to disease severity variation. We derived diverse gut communities by colonizing germ-free mice with human fecal communities. The mice were then infected with a single *C. difficile* RT027 isolate which resulted in mortality and histopathologic differences. The severity grouped by the human fecal community they received. Generally, facultative anaerobes with pathogenic potential, such as *Escherichia*, *Helicobacter*,  and *Klebsiella*, were associated with more severe disease. Bacterial groups associated with dietary fiber degradation, such as *Coprobacillus*, were associated with reduced disease severity. Lastly, we showed communities that resulted in either low or high histopathologic scores or severe disease had the same outcome when infected with a different *C. difficile* isolate.

@@

*Clostridioides difficile* infection (CDI) can be asymptomatic, or range in severity from mild diarrhea to severe outcomes of surgical interventions and even death. Models that predict severity and guide treatment decisions are based on clinical factors and *C. difficile* characteristics. The gut microbiome plays a role in potecting against CDI, though has not been investigated for its effect on CDI disease severity. We demonstrate that variation in the microbiome of mice colonized with human feces was sufficient to result in a range of disease severities. The results reveal groups of bacteria associated with both severe and more mild *C. difficile* infections. Gut bacterial community data from patients with CDI could improve our ability to identify patients at risk for severe disease and improve interventions which treat the gut bacteria to reduce host damage.

@@

*Clostridioides difficile* infections (CDI) have been increasing in incidence and severity since *C. difficile* was first identified as the cause of antibiotic-associated psuedomembranous colitis [@Kelly2008]. Disease severity can range from asymptomatic to mild diarrhea to toxic megacolon and death. Treatment for CDI is driven by the disease severity. The Infectious Diseases Society of America (IDSA) and Society for Healthcare Epidemiology of America (SHEA) guidelines categorize severe disease based on white blood cell count >15,000 cells/mm^3 and/or a serum creatinine >1.5 mg/dL, and complicated disease based on systolic blood pressure <90, ileus or toxic megacolon [@McDonald2018]. Since these measures are only present once the disease has advanced, the IDSA classification does not allow for identification of patients at risk of severe CDI. Thus, we have limited ability to prevent patients from developing severe CDI. Ideally, we could identify patients that are high-risk of severe disease use that to adjust their treatments to halt the disease before it worsens.

Prediction models have been developed to score a patients risk for severe CDI outcomes but have not been robust for broad application [@Dieterle2020; @Butt2013; @Perry2021; @vanBeurden2017]. Firstly, there is no widely accepted definition of severe disease [@Ressler2021]. While the IDSA guideline is frequently used, many studies develop their own severity definitions. Aside from the features utilized by the IDSA guidelines, other severity measures use patient demographics, vital information, laboratory test results, disease complications, or infection outcomes such as ICU admission, surgery and/or death [@Dieterle2020; @Zar2007; @Gujja2009; @Jardin2013]. However, a recent multi-center external validation of these scoring systems using a consistent severity definition - transfer or admission to the intensive care unit, colectomy, and/or death within 30 days of diagnosis - demonstrated a decline in performance and high false positivity rate [@Perry2021]. Thus, it may be necessary to incorporate information in addition to patient and clinical factors to accurately idenitfy patients at risk of severe CDI.

CDI disease severity is the result of damage to the host. Severity has been associated with characteristics of *C. difficile* ribotypes, such as 027, and host factors [@Bauer2011]. Ribotypes with genes for toxin TcdA/TcdB have the potential to disrupt the tight junctions between the intestinal epithelial cells [@Abt2016]. The presence of binary toxin CDT or mutations in *tcdC*, the negative regulator of toxin transcription, have been associated with increased virulence of TcdA/TcdB [@Abt2016]. In addition to the pathogenic potential of *C. difficile*, the response by the host also affects the disease severity. IgG antibody activity towards the TcdA/TcdB are decreased in both patients with symptomatic and recurrent infections [@Kelly2011]. In addition to controlling the infection, the immune response can also cause intestinal injury. Increased neutrophil inflitration into the mucosa after the cellular tight junctions have been compromised can exacerbate the inflammatory response [@Farooq2015]. While we have have elucidated these major drivers of host damage in CDI, these do not explain the observed variation in disease severity.

Missing from our CDI disease model are the effects of gut bacteria. *C. difficile* interacts with the intestinal community in many ways. The indigenous bacteria of a healthy intestinal community provide a protective barrier preventing *C. difficile* from infecting the gut. A range of mechanisms can disrupt this barrier, including antibiotics, medications, or dietary changes, and lead to increased susceptiblity to CDI [@Britton2014; @Hryckowian2018; @VichVila2020]. Once in the gut, the indigenous bacteria can either promote or inhibit *C. difficile* through producing molecules or modifying the environment [@Abbas2020]. Bile acids metabolized by the gut bacteria can inhibit *C. difficile* growth and affect toxin production [@Sorg2008; @Thanissery2017]. Bacteria in the gut also can compete more directly with *C. difficile* through antibiotic production or nutrient consumption [@Aguirre2021; @Kang2019]. While the interaction between the gut bacteria and *C. difficile* is established, the effect the gut bacteria on CDI disease severity is unclear. Recent studies that have begun to uncover this relationship but have either used clinical data which does not permit control of the host or *C. difficile* variation [@Seekatz2016] or murine experiments which lack variation in the bacterial communities [@Dieterle2020]. Recently, Tomkovich et al. showed the variation in the bacterial communities between mice from different mouse colonies and vendors resulted in different clearance rates of *C. difficile* [@Tomkovich2020]. Therefore it is clear the gut bacteria modulate the activity and infection dynmaics of *C. difficile*, we hypothesized that gut bacteria also contribute to the variation in CDI disease severity. Here, we colonized germ-free C57BL/6 mice with human fecal communities to test if the variablity in the gut bacteria can produce and explain variation in the CDI disease severity.

@@

***C. difficile* is able to infect germ-free mice colonized with human fecal microbial communities without perturbation.**  Based on our previous observation that variation in the murine microbiota associated with different *C. difficile* colonization dynamics [@Tomkovich2020], we sought to explore how variation in murine microbial communities affects the outcome of *C. difficile* infection. To produce communities with greater variation than mouse colonies with their indigenous communties, we used human fecal communities to establish a microbial community in the intestine of germ-free mice [@Nagpal2018]. We inoculated germ-free C57BL/6 mice with homogenized feces from one of 15 human fecal samples via oral gavage. The human fecal samples selected were chosen to provide diverse communtiy structures and based on community clusters [@Schubert2014]. The gut community was allowed to establish itself for two weeks post inoculation [@Gillilland2012]. We then surveyed the bacterial members of the gut communities by 16S rRNA gene sequencing of fecal pellets (Figure 1A). The bacterial communities from each mouse grouped more closely to those communities from mice that received the same human fecal donor community than to either their respective donor community or any mice who received a fecal homogenate from a different donor (Figure 1B). Most of the communities were primairily composed of populations of *Clostridia*, *Bacteroidia*, *Erysipelotrichia*, *Bacilli*, and *Gammaproteobacteria*. However, each group of mice harbored unique combination of class abundances in their gut bacterial communities.

Next, we tested this set of mice with their human-derived gut microbial communities for susceptibility to *C. difficile* infection. A typical mouse model of CDI requires pretreatment with antibiotics such as clindamycin to become susceptible to *C. difficile* colonization [@Chen2008; @Schubert2015]. However, we wanted to avoid modifying the communties with an antibiotic to maintain their unique microbial compositions and ecological relationships. Since these communties came from people who have increased risk of CDI, such as hospitalization or recent antibiotic use [@Schubert2014], or the communities became susceptible through the murine gut colonization process, we decided to test if *C. difficile* was able to infect these mice without any antibiotic perturbation. We hypothesized that *C. difficile* would be able to colonize a subset of mice whose communities were susceptible from perturbations either the donor experienced or from the ecological selection colonizing the murine gut. After two weeks post incoluation with human fecal communities, the mice were then challenged with $10^{3}$ *C. difficile* ribotype 027 human isolate 431 spores. The mice were followed for 10 days post-challenge and their stool was collected and plated for *C. difficile* colony forming units (CFU) to detect infection level. Suprisingly, communities from all donors were colonized (Figure 2). Only two mice were able to resist *C. difficile* infection and both received their community from the same donor, Donor C. The remaining mice demonstrated susceptiblity to *C. difficile* infection without the need for any perturbation. Thus, *C. difficile* infected the mice without needing to perturb the established human-derived gut microbial communities, allowing them to maintain their unique community structures.

**Infection severity varies by initial community.** After challenging the mice with *C. difficile*, we investigated the outcome from the *C. difficile* infection and its relationship to the initial community. We followed the mice for 10 days post-challenge for *C. difficile* CFU, toxin production, and mortality. Seven mice, from Donors A, B and E, were not initally colonized on the day after *C. difficile* challenge but were infected ($>10^{6}$) by the end of the experiment. All mice that received their community from Donor J through O suffered from severe disease, which became highly colonized one day post-challenge and moribund within 3 days post-challenge. The remaining mice, except the Donor C mice which were not infected, maintained *C. difficile* infection through the end of the experiment (Figure 2). At 10 days post-challenge, or earlier for the moribund mice, mice were euthanised and cecal tissue was collected and scored for histopathologic signs of disease and fecal pellets were assayed for toxin activity (Figure 3). Overall, there was greater toxin activity detected in the stool of the moribund mice (*P* = 0.003). However, when looking at each group of mice, we saw there was a range in toxin activity for both the moribund and non-moribund mice (Figure 3A). Non-moribund mice from Donors D through I had comparable toxin activity as the moribund mice. Additionally, not all moribund mice had toxin activity detected in their stool. This discordance between toxin activity and disease severity has been previously described [@Akerlund2006; @Vitucci2020]. Next we examined the cecal tissue for histopathologic damage. For moribund mice, we observed high levels of epithelial damage, tissue edema and inflammation (Figure S1), similar to previously reported histopathologic findings for *C. difficile* ribotype 027 [@Cowardin2016]. As observed with toxin activity, the moribund mice had higher histopathologic scores than the mice with non-severe disease (*P* = 3.0e-9). However, unlike the toxin activity, all moribund mice had consistently high histopathologic summary scores (Figure 3B). For Donor groups A through I, where all mice were non-moribund, we saw a range in tissue damage from no signs of disease to nearly the same level as the mice with severe disease, which grouped by community donor. Together, the toxin activity, histopathologic score, and moribundity showed variation across the donors but were largely consistent within each group of mice recieving the same human donor fecal community.

**Microbial community members explain variation in CDI severity.** Since the disease charateristics grouped by human donor fecal community, we next investigated specific bacterial populations that could explain the variation in disease severity. To this aim, we interogated the day 0 bacterial community for their relationship to our CDI severity measurements using linear discriminant analysis (LDA) effect size (LEfSe) analysis. We split the mice into groups by severity level based on their moribundity and histopathologic score. We dichotomized the histopathologic scores into high and low bins by splitting on the median score. This analysis revealed 20 genera that were significantly different by the disease severity (Figure 4A). Bacterial genera *Turicibacter*, *Streptococcus*, *Staphylococcus*, *Pseudomonas*, *Phocaeicola*, *Parabacteroides*, *Bacteroides*, and *Escherichia/Shigella* were detected at higher relative abundances in the mice that became moribund. Populations of *Anaerotignum*, *Coprobacillus*, *Enterocloster*, and *Murimonas* were more abundant in the mice that would develop only low intestinal injury. To understand the role of toxin activity in disease severity, we applied LEfSe to identify the genera most likely to explain the observed differences in toxin activity (Figure 4B). Many genera the were associated with toxin production as we would expect from their association with moribundity, such as populations of *Escherichia/Shigella* and *Bacteroides*. Likewise, there were genera such as *Anaerotignum*, *Enterocloster*, and *Murimonas* that exhibited greater relative abundance in communities from non-moribund mice with a low histopathologic score and no detected toxin. Overall, communities without genera that demonstrate a linear trend between relative abundance and disease severity can be better understood by their association with toxin activity. Lastly, we tested for correlations between the histopathologic score and the endpoint relative abundances of bacterial operational taxonomic units (OTUs) (Figure 4C). The endpoint relative abundance of *Bacteroides* was positively correlated with histopathologic score, as its day 0 relative abundance did with disease severity (Figure 4A). Populations of *Klebsiella* and *Prevotellaceae* were positively correlated with the histopathologic score and were increased in the group of mice with detectable toxin. Our analysis have revealed bacterial taxa that are associated with three different measures of CDI severity in our model: moribundity, histopathologic score, and toxin activity.

We next wanted to determine whether collectively, bacterial community membership and relative abundance could be predictive of CDI severity. We trained random forest models with the day 0 bacterial community relative abundance data at each taxonomic rank to predict toxin production, severe disease or final histopathologic score. Overall for predicting toxin production, microbial populations aggregated by their phylum level classificiation performed as well as models using lower taxonomic ranks (AUROC = 0.83, Figure S2). Collectively, members of *Verrucomicrobia*, *Campilobacterota* and *Proteobacteria* contributed the most to the correct prediction of toxin production (Figure 5A). *C. difficile* was more likely to produce toxin when the community it infected had less abundant populations of *Verrucomicrobia* and *Campilobacterota* and had more abundant populations of *Proteobacteria*. Next, we assessed the ability of the community to predict moribundity. Class rank classification was sufficient to predict if the mouse would succumb to the infection before the end of the experiment (AUROC = 0.91, Figure S2). Bacterial taxa belonging to *Bacilli*, *Firmicutes*, and *Erysipelotrichia* hasd the greatest effect on model accuracy. Larger populations of *Bacilli* and *Firmicutes* and reduced populations of *Erysipelotrichia* were more likely to result in moribundity (Figure 5B). The only other class that was decreased in moribund mice was *Clostridia*. Lastly, the relative abundances of genera were able to predict a high or low histopathologic score (histopathologic score dichotomized as in previous analysis, AUROC = 0.99, Figure S2). No genera had a significantly greater effect on the model performance than any others, indicating the model is reliant on many genera for the correct prediction. The model used some of the genera identified in the LEfSe analysis, such as *Coprobacillus*, *Anaerostipes*, and *Hungatella*. Communities with greater abundances of *Hungatella*, *Eggerthella*, *Bifidobacterium*, *Duncaniella* and *Neisseria* were more likely to have high histopathologic scores.

**Disease severity consistent by donor community across isolates.** Since Carlson et al. showed clinical disease severity varied by *C. difficile* isolate characteristics [@Carlson2013; @Carlson2015], we next wanted to test if the effect of the community on CDI severity is conserved across *C. difficile* isolates. To facilitate this analysis, we selected a subset of donor fecal samples based on their associated mouse transplant communities that fit the following criteria: one community we would expect to result in a low histopathologic score (< 5), a high histopathologic score (> 5), and a high histopathologic score which becomes moribund. Additionally, we selected the donors based their relative abundances of genera identified through the our analysis (Figure S3). We inoculated mice via oral gavage of the human fecal slurry of either Donor F, G, or M and allowed two weeks for the a community to establish. We then challenged the mice with a different *C. difficile* isolate and all mice were infected. The moribundity and histopathologic score outcomes from challenge with the second *C. difficile* human isolate were comparable to the outcomes of challenges with *C. difficile* ribotype 027 human isolate 431 by their donor group (Figure 6).

@@

Challenging mice colonized with different human fecal communities with a single *C. difficile* isolate allowed us to identify the effect of microbiome variation on *C. difficile* infection disease severity. Our LEfSe analyses and random forest models revealed the relationships between the bacterial community members and the degree of disease severity. We used the relative abundance patterns of community members with strong associations with disease severity to select communities to test if the observed disease severity is conserved when a second *C. difficile* isolate is used. The disease outcome from the *C. difficile* chalenge with the new isolate matched the expected disease severities. Therefore, the effect of the community variation on disease severity *in vivo* is greater than the effect of *C. difficile* isolate phenotype variation [@Carlson2013; @Carlson2015]. Overall, these results have demonstrated that the bacterial community contributes to the disease severity of *C. difficile* infection.

CDI severity can range from no symptoms, mild symptoms such as diarrhea, or severe symptoms such as toxic megacolon or death. Physicians use prediction tools to identify patients most at risk of developing a severe infection using white blood cell counts, serum albumin level, or serum creatine level [@Lungulescu2011; @McDonald2018; @Zar2007]. Those levels are being driven by the activities in the intestine [@diMasi2018]. Research into the drivers of this variation have revealed factors that make *C. difficile* more virulent. Strains are categorized for their virulence by the presence and production toxin A, toxin B, and binary toxin and the prevelance in outbreaks, such as ribotype 027 and 078 [@AbernathyClose2020; @Theriot2011; @Goorhuis2008; @OConnor2009; @Chen2008]. However, other studies have shown that virulence is not necessarily linked with toxin production [@Rao2015] or the strain [@Walk2012]. Furthermore, there is variation in the genome, growth rate, sporulation, germination, and toxin production in different isolates of a strain [@Carlson2013; @He2010]. This variation may help explain why severe CDI prediction tools often miss identifying many patients with CDI that will develop severe disease [@Dieterle2020; @Butt2013; @Perry2021; @vanBeurden2017]. Therefore, it is necessary to gain a full understanding of all factors contributing to disease variation to improve our ability to predict severity.
